---
title: "HNC Tumour Marker Identification Pipeline"
subtitle: "TCGA-HNSC & GSE65858 | All HNC Sites | HPV-Negative Only | Tumour vs. Normal"
author: "Khushi"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 7
)
```

---

# 0. Install & Load Packages

```{r packages}
# Install if needed (run once, then comment out)
# if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# BiocManager::install(c("TCGAbiolinks", "DESeq2", "limma", "edgeR", "GEOquery"))
# install.packages(c("tidyverse", "pheatmap", "ggrepel", "RColorBrewer", "openxlsx"))

library(TCGAbiolinks)   # TCGA data download
library(GEOquery)       # GEO data download
library(DESeq2)         # TCGA differential expression (count data)
library(limma)          # GSE65858 differential expression (microarray)
library(edgeR)          # Normalisation support
library(tidyverse)      # Data wrangling
library(pheatmap)       # Heatmaps
library(ggrepel)        # Volcano plot labels
library(RColorBrewer)   # Colour palettes
library(openxlsx)       # Export results to Excel
```

---

# 1. TCGA-HNSC Analysis

## 1.1 Download TCGA-HNSC RNA-seq Data

```{r tcga-download, eval=FALSE}
# Query TCGA-HNSC RNA-seq (HTSeq counts)
query_tcga <- GDCquery(
  project           = "TCGA-HNSC",
  data.category     = "Transcriptome Profiling",
  data.type         = "Gene Expression Quantification",
  workflow.type     = "STAR - Counts",
  sample.type       = c("Primary Tumor", "Solid Tissue Normal")
)

GDCdownload(query_tcga)
tcga_data <- GDCprepare(query_tcga)

# Save to avoid re-downloading
saveRDS(tcga_data, "tcga_hnsc_raw.rds")
```

```{r tcga-load}
# Load saved data (use this after first download)
tcga_data <- readRDS("tcga_hnsc_raw.rds")
```

## 1.2 Extract Counts & Clinical Metadata

```{r tcga-prep}
# Extract count matrix and sample metadata
tcga_counts <- assay(tcga_data, "unstranded")
tcga_meta   <- as.data.frame(colData(tcga_data))

# Classify tumour vs normal
tcga_meta$condition <- ifelse(
  tcga_meta$sample_type == "Solid Tissue Normal",
  "Normal", "Tumour"
)

# ----- HPV FILTERING -----
# TCGA-HNSC uses 'paper_HPV.status' (values: "Positive", "Negative", NA)
# Normal tissue samples do not have HPV status assigned — we retain ALL normals
# since they are HPV-agnostic reference tissue

cat("HPV status breakdown (tumour samples only):\n")
print(table(tcga_meta$paper_HPV.status, useNA = "ifany"))

# Keep: HPV-negative tumours + all normal tissue samples
tcga_meta <- tcga_meta %>%
  filter(
    (condition == "Tumour" & paper_HPV.status == "Negative") |
    condition == "Normal"
  )

# Subset count matrix to match filtered samples
tcga_counts <- tcga_counts[, rownames(tcga_meta)]

cat("\nSamples after HPV-negative filtering:\n")
table(tcga_meta$condition)
```

## 1.3 Pre-filtering

```{r tcga-filter}
# Remove lowly expressed genes (keep genes with ≥10 counts in ≥10 samples)
keep <- rowSums(tcga_counts >= 10) >= 10
tcga_counts_filt <- tcga_counts[keep, ]

cat("Genes before filtering:", nrow(tcga_counts), "\n")
cat("Genes after filtering: ", nrow(tcga_counts_filt), "\n")
```

## 1.4 DESeq2 Differential Expression

```{r tcga-deseq2}
# Build DESeq2 object
dds <- DESeqDataSetFromMatrix(
  countData = tcga_counts_filt,
  colData   = tcga_meta,
  design    = ~ condition
)

# Set reference level to Normal
dds$condition <- relevel(dds$condition, ref = "Normal")

# Run DESeq2
dds <- DESeq(dds)

# Extract results: Tumour vs Normal
tcga_res <- results(dds,
  contrast  = c("condition", "Tumour", "Normal"),
  alpha     = 0.05
)

# Convert to data frame and clean up
tcga_res_df <- as.data.frame(tcga_res) %>%
  rownames_to_column("gene") %>%
  filter(!is.na(padj)) %>%
  arrange(padj, desc(log2FoldChange))

# Summary
cat("Significantly upregulated in Tumour (padj < 0.05, log2FC > 1):",
    sum(tcga_res_df$padj < 0.05 & tcga_res_df$log2FoldChange > 1, na.rm = TRUE), "\n")
```

## 1.5 TCGA Volcano Plot

```{r tcga-volcano}
# Label top 30 upregulated genes
tcga_res_df <- tcga_res_df %>%
  mutate(
    sig     = case_when(
      padj < 0.05 & log2FoldChange >  1 ~ "Upregulated",
      padj < 0.05 & log2FoldChange < -1 ~ "Downregulated",
      TRUE ~ "NS"
    ),
    neg_log10_p = -log10(padj)
  )

top30_tcga <- tcga_res_df %>%
  filter(sig == "Upregulated") %>%
  slice_max(log2FoldChange, n = 30)

ggplot(tcga_res_df, aes(x = log2FoldChange, y = neg_log10_p, colour = sig)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_label_repel(
    data    = top30_tcga,
    aes(label = gene),
    size    = 2.8,
    max.overlaps = 20
  ) +
  scale_colour_manual(values = c(
    "Upregulated"   = "#C0392B",
    "Downregulated" = "#2980B9",
    "NS"            = "grey70"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  labs(
    title    = "TCGA-HNSC: Tumour vs. Normal",
    x        = "log2 Fold Change",
    y        = "-log10(adjusted p-value)",
    colour   = NULL
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## 1.6 TCGA Top 30 Upregulated Markers

```{r tcga-top30}
tcga_top30 <- tcga_res_df %>%
  filter(sig == "Upregulated") %>%
  slice_max(log2FoldChange, n = 30) %>%
  select(gene, log2FoldChange, padj, baseMean)

knitr::kable(tcga_top30, digits = 4, caption = "TCGA-HNSC: Top 30 Upregulated Genes")
```

## 1.7 TCGA Heatmap (Top 30)

```{r tcga-heatmap}
# Extract normalised counts for top 30 genes
tcga_vst      <- vst(dds, blind = FALSE)
top30_genes   <- tcga_top30$gene
heatmap_mat   <- assay(tcga_vst)[top30_genes, ]

# Annotation: Tumour vs Normal
anno_col <- data.frame(
  Condition = tcga_meta$condition,
  row.names = colnames(heatmap_mat)
)

pheatmap(
  heatmap_mat,
  annotation_col   = anno_col,
  annotation_colors = list(Condition = c(Tumour = "#C0392B", Normal = "#2980B9")),
  show_colnames    = FALSE,
  scale            = "row",
  color            = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  main             = "TCGA-HNSC: Top 30 Upregulated Markers",
  fontsize_row     = 9
)
```

## 1.8 TCGA Boxplots (Top 10)

```{r tcga-boxplots, fig.height=12}
# Normalised counts for top 10 genes
top10_tcga   <- tcga_top30$gene[1:10]
vst_long     <- assay(tcga_vst)[top10_tcga, ] %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "expression") %>%
  left_join(tcga_meta %>% select(barcode, condition) %>%
              rename(sample = barcode), by = "sample")

ggplot(vst_long, aes(x = condition, y = expression, fill = condition)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.8) +
  facet_wrap(~ gene, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c(Tumour = "#C0392B", Normal = "#2980B9")) +
  labs(
    title = "TCGA-HNSC: Top 10 Marker Expression",
    x     = NULL, y = "VST-normalised expression"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")
```

---

# 2. GSE65858 Analysis

## 2.1 Download GSE65858

```{r gse-download, eval=FALSE}
gse <- getGEO("GSE65858", GSEMatrix = TRUE, destdir = ".")
saveRDS(gse, "gse65858_raw.rds")
```

```{r gse-load}
gse      <- readRDS("gse65858_raw.rds")
gse_eset <- gse[[1]]   # ExpressionSet object
```

## 2.2 Extract Expression & Metadata

```{r gse-prep}
gse_expr <- exprs(gse_eset)           # Expression matrix (already normalised)
gse_meta <- pData(gse_eset)           # Sample metadata

# Check available columns for tumour/normal classification
# NOTE: Inspect gse_meta column names to confirm the correct field
# The line below may need adjusting based on actual column names
cat("Metadata columns:\n")
print(colnames(gse_meta))
```

```{r gse-classify}
# Inspect metadata to find HPV status field
# GSE65858 typically stores this in characteristics_ch1 or a dedicated column
# Print first few rows to confirm field name
print(head(gse_meta[, grepl("hpv|status|charact", colnames(gse_meta), ignore.case = TRUE)]))

# Classify condition (tumour vs normal)
gse_meta$condition <- ifelse(
  grepl("normal|control", gse_meta$source_name_ch1, ignore.case = TRUE),
  "Normal", "Tumour"
)

# ----- HPV FILTERING -----
# GSE65858 stores HPV status in characteristics — adjust field name below
# after inspecting the output above. Common field: "characteristics_ch1.1" or similar
# Example extraction (update column name as needed):
gse_meta$hpv_status <- gsub(".*hpv status: ", "", gse_meta$`characteristics_ch1.1`,
                              ignore.case = TRUE)

cat("HPV status breakdown (tumour samples only):\n")
print(table(gse_meta$hpv_status[gse_meta$condition == "Tumour"], useNA = "ifany"))

# Keep: HPV-negative tumours + all normal samples
gse_meta <- gse_meta %>%
  filter(
    (condition == "Tumour" & grepl("negative|neg|0", hpv_status, ignore.case = TRUE)) |
    condition == "Normal"
  )

# Subset expression matrix to match filtered samples
gse_expr <- gse_expr[, rownames(gse_meta)]

cat("\nSamples after HPV-negative filtering:\n")
table(gse_meta$condition)
```

## 2.3 limma Differential Expression

```{r gse-limma}
# Design matrix
condition      <- factor(gse_meta$condition, levels = c("Normal", "Tumour"))
design         <- model.matrix(~ condition)

# Fit linear model
fit  <- lmFit(gse_expr, design)
fit  <- eBayes(fit)

# Extract results
gse_res <- topTable(fit,
  coef      = "conditionTumour",
  number    = Inf,
  adjust    = "BH",
  sort.by   = "logFC"
) %>%
  rownames_to_column("gene") %>%
  filter(!is.na(adj.P.Val)) %>%
  arrange(adj.P.Val, desc(logFC))

cat("Significantly upregulated in Tumour (adj.P.Val < 0.05, logFC > 1):",
    sum(gse_res$adj.P.Val < 0.05 & gse_res$logFC > 1, na.rm = TRUE), "\n")
```

## 2.4 GSE65858 Volcano Plot

```{r gse-volcano}
gse_res <- gse_res %>%
  mutate(
    sig = case_when(
      adj.P.Val < 0.05 & logFC >  1 ~ "Upregulated",
      adj.P.Val < 0.05 & logFC < -1 ~ "Downregulated",
      TRUE ~ "NS"
    ),
    neg_log10_p = -log10(adj.P.Val)
  )

top30_gse <- gse_res %>%
  filter(sig == "Upregulated") %>%
  slice_max(logFC, n = 30)

ggplot(gse_res, aes(x = logFC, y = neg_log10_p, colour = sig)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_label_repel(
    data    = top30_gse,
    aes(label = gene),
    size    = 2.8,
    max.overlaps = 20
  ) +
  scale_colour_manual(values = c(
    "Upregulated"   = "#C0392B",
    "Downregulated" = "#2980B9",
    "NS"            = "grey70"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  labs(
    title  = "GSE65858: Tumour vs. Normal",
    x      = "log2 Fold Change",
    y      = "-log10(adjusted p-value)",
    colour = NULL
  ) +
  theme_bw(base_size = 13) +
  theme(legend.position = "top")
```

## 2.5 GSE65858 Top 30 Upregulated Markers

```{r gse-top30}
gse_top30 <- gse_res %>%
  filter(sig == "Upregulated") %>%
  slice_max(logFC, n = 30) %>%
  select(gene, logFC, adj.P.Val, AveExpr)

knitr::kable(gse_top30, digits = 4, caption = "GSE65858: Top 30 Upregulated Genes")
```

## 2.6 GSE65858 Heatmap (Top 30)

```{r gse-heatmap}
top30_gse_genes <- gse_top30$gene
heatmap_gse     <- gse_expr[top30_gse_genes, ]

anno_gse <- data.frame(
  Condition = gse_meta$condition,
  row.names = colnames(heatmap_gse)
)

pheatmap(
  heatmap_gse,
  annotation_col    = anno_gse,
  annotation_colors = list(Condition = c(Tumour = "#C0392B", Normal = "#2980B9")),
  show_colnames     = FALSE,
  scale             = "row",
  color             = colorRampPalette(c("#2980B9", "white", "#C0392B"))(100),
  main              = "GSE65858: Top 30 Upregulated Markers",
  fontsize_row      = 9
)
```

## 2.7 GSE65858 Boxplots (Top 10)

```{r gse-boxplots, fig.height=12}
top10_gse <- gse_top30$gene[1:10]

gse_long <- gse_expr[top10_gse, ] %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "expression") %>%
  left_join(gse_meta %>% rownames_to_column("sample") %>%
              select(sample, condition), by = "sample")

ggplot(gse_long, aes(x = condition, y = expression, fill = condition)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.8) +
  facet_wrap(~ gene, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c(Tumour = "#C0392B", Normal = "#2980B9")) +
  labs(
    title = "GSE65858: Top 10 Marker Expression",
    x     = NULL, y = "Normalised expression"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "none")
```

---

# 3. Consensus Marker Identification

## 3.1 Overlap Between Both Datasets

```{r consensus}
tcga_up_genes <- tcga_top30$gene
gse_up_genes  <- gse_top30$gene

# Consensus: genes upregulated in BOTH datasets
consensus_genes <- intersect(tcga_up_genes, gse_up_genes)

cat("Top 30 upregulated in TCGA:   ", length(tcga_up_genes), "genes\n")
cat("Top 30 upregulated in GSE65858:", length(gse_up_genes), "genes\n")
cat("Consensus (in both):           ", length(consensus_genes), "genes\n")
print(consensus_genes)
```

## 3.2 Consensus Marker Table

```{r consensus-table}
# Combine stats from both datasets for consensus genes
consensus_df <- tcga_top
